---
title: "Biostat M280 Homework 3"
subtitle: Due Mar 1 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Use tidyverse and bash to explore following two data sets.

## Q1 LA City Employee Payroll

The `/home/m280-data/la_payroll/City_Employee_Payroll.csv` file on teaching server contains payroll information of LA City employees in years 2013-2018. It was downloaded from [LA City Controller's Office](https://controllerdata.lacity.org/Payroll/City-Employee-Payroll/pazn-qyym). Make a Shiny app to facilitate visualization of this data. 

1. For efficiency of the Shiny app, you should first pre-process, pare down, tidy, and save the data, e.g., as a compressed RDS file, to be used in the app.

```{bash}
ln -sf /home/m280data/la_payroll/City_Employee_Payroll.csv City_Employee_Payroll.csv
```

0. **Total payroll by LA City**. Visualize the total LA City payroll of each year, with breakdown into base pay, overtime pay, and other pay.

```{r}
library("rio")
library("dplyr")
library("magrittr")
library(reshape2)
options(scipen = 999)
employee <- read.csv("/home/m280data/la_payroll/City_Employee_Payroll.csv", 
                     stringsAsFactors = FALSE, header = TRUE)
(employee %<>% select(Year, Department.Title, Job.Class.Title, Total.Payments, 
                      Base.Pay, Overtime.Pay))
employee$Other.Pay <- employee$Total.Payments - employee$Base.Pay - employee$Overtime.Pay
employee.long <- melt(data, id.vars = c("Year", "Department.Title", "Job.Class.Title"))
saveRDS(employee, file = "City_Employee_Payroll.rds")
saveRDS(employee.long, file = "City_Employee_Payroll_Long.rds")
```

0. **Who earned most?** Visualize the payroll information (total payment with breakdown into base pay, overtime pay, and other pay, Department, Job Title) of the top $n$ highest paid LA City employees in a specific year. User specifies $n$ (default 10) and year (default 2017).

0. **Which departments earn most?** Visualize the mean or median payroll, with breakdown into base pay, overtime pay, and other pay, of top $n$ earning departments. User specifies $n$ (default 5), year (default 2018), and method (mean or median, default median).

0. **Which departments cost most?** Visualize the total payroll, with breakdown into base pay, overtime pay, and other pay, of top $n$ expensive departments. User specifies $n$ (default 5) and year (default 2017).

0. Visualize any other information you are interested in.

0. Publish your Shiny app to <https://www.shinyapps.io> and share the link.

## Q2 LA City Parking War

The SQLite database `/home/m280data/la_parking/LA_Parking_Citations.sqlite` on teaching server contains information about parking tickets in LA City. It was downloaded from [LA Open Data Portal](https://data.lacity.org/A-Well-Run-City/Parking-Citations/wjz9-h9np). Connect to the database and answer following questions using plots and summary statistics. In this exercise, you are **not** allowed to load whole data into memory. Use the _transform in database, plot in R_ strategy.

```{r}
# creating database
library("DBI")
library("RSQLite")
if (Sys.info()[["sysname"]] == "Linux") {
  db <- dbConnect(RSQLite::SQLite(), 
                  dbname = "/home/m280data/la_parking/LA_Parking_Citations.sqlite")
} else if (Sys.info()[["sysname"]] == "Darwin") {
  db <- dbConnect(RSQLite::SQLite(), 
                  dbname = "./LA_Parking_Citations.sqlite")
}

dbListTables(db)
```

1. How many tickets are in this data set? Which time period do these tickets span? Which years have most data?

```{r}
# creating object from database
la_sql <- dplyr::tbl(db, "latix")
ticket <- la_sql %>% summarize(n = n())
year <- la_sql %>%
  group_by(Issue_Year) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
ticket
year
range <- la_sql %>%
  group_by(Issue_Year, Issue_Month, Issue_Day) %>%
  summarize(n = n())
range
```
There are 7,656,418 tickets in this dataset. The range is 2015-07-02 to 2018-03-26. The year with the most tickets is 2017 with 2,235,243 entries. The year with the 2nd most is 2016 with 2,177,956 entries.

0. When (which hour, weekday, month day, and month) are you most likely to get a ticket and when are you least likely to get a ticket?

```{r}
hour <- la_sql %>%
  group_by(Issue_Hour, Issue_Minute) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
weekday <- la_sql %>%
  group_by(Issue_Wday) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
date <- la_sql %>%
  group_by(Issue_Day) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
month <- la_sql %>%
  group_by(Issue_Month) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
month
date
weekday
hour
```
The time one is most likely to be ticketed is 12:10, weekday 3 (Wednesday?), 13th of the month, August. The least likely time is 22:37 (or 21:13), weekday 1 (Monday?, excluding NA entries), 31st of the month (excluding NA entries), February (excluding NA entries).

0. Which car makes received most citations?

```{r}
maker <- la_sql %>%
  group_by(Make) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
maker
```
TOYT/Toyota received the most citations at 1,270,804 entries.

0. How many different colors of cars were ticketed? Which color attracted most tickets?

```{r}
color <- la_sql %>%
  group_by(Color) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
color
```
BK (Black?) attracted the most tickets at 1,652,621 entries. While there are too many unique answers presumably due to typos/variations of a single answer, some common colors also include WT (white?), GY (grey?), SL (silver?) and BL (blue?).

0. What are the most common ticket types?

```{r}
head(la_sql)
type <- la_sql %>%
  group_by(Violation_Code) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
type
```
The most common ticket types are violation codes 80.69BS and 88.13B+ at 2,166,602 and 1,302,155 entries respectively. Codes 80.56E4+ (547,904 entries) and 80.58L (523,775 entries) were also common.

0. How much money was collected on parking tickets in 2016, 2017 and 2018?
```{r}
m2016 <- la_sql %>%
  filter(Issue_Year == 2016) %>%
  summarize(n = sum(Fine_amount), na.rm = TRUE)
m2017 <- la_sql %>%
  filter(Issue_Year == 2017) %>%
  summarize(n = sum(Fine_amount), na.rm = TRUE)
m2018 <- la_sql %>%
  filter(Issue_Year == 2018) %>%
  summarize(n = sum(Fine_amount), na.rm = TRUE)
m2016
m2017
m2018
```
$152,145,538 was collected in 2016. $157,122,489 was collected in 2017. $138,875,787 was collected in 2018.

0. If you've been ticketed in LA County, did you find your ticket in this data set?

No, I have not been ticketed in LA County.

0. Read the blog <http://www.brettrics.com/9-million-parking-tickets-la/> and try to reproduce plots using your data.

```{r}
library(ggplot2)
library(tidyverse)
head(la_sql)
la_sql %>%
  mutate(Issue_Date = paste(Issue_Year, Issue_Month)) %>%
  group_by(Issue_Date) %>%
  summarize(n = n()) %>%
  ggplot(aes(x = Issue_Date, y = n, group = 1)) + 
    geom_line() +
    theme(axis.text.x = element_text(angle=45),
          axis.text.y = element_text(angle=45))
```
