---
title: "Biostat M280 Homework 3"
subtitle: Due Mar 1 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Use tidyverse and bash to explore following two data sets.

## Q1 LA City Employee Payroll

The `/home/m280-data/la_payroll/City_Employee_Payroll.csv` file on teaching server contains payroll information of LA City employees in years 2013-2018. It was downloaded from [LA City Controller's Office](https://controllerdata.lacity.org/Payroll/City-Employee-Payroll/pazn-qyym). Make a Shiny app to facilitate visualization of this data. 

1. For efficiency of the Shiny app, you should first pre-process, pare down, tidy, and save the data, e.g., as a compressed RDS file, to be used in the app.

```{bash}
ln -sf /home/m280-data/la_payroll/City_Employee_Payroll.csv City_Employee_Payroll.csv
```

0. **Total payroll by LA City**. Visualize the total LA City payroll of each year, with breakdown into base pay, overtime pay, and other pay.

```{r}
library(shiny)
employee <- convert("City_Employee_Payroll.csv", "City_Employee_Payroll.rds")
ui <- fluidPage(

  # app title
  titlePanel("Payroll Visualization"),

  # sidebar with input/output code
  sidebarLayout(

    # sidebar, has input code
    sidebarPanel(

      # Input: Selector for choosing dataset ---- USE FOR PAY TYPE?
      selectInput(inputId = "dataset",
                  label = "Choose a dataset:",
                  choices = c("rock", "pressure", "cars")),

      # Input: Numeric entry for number of obs to view ---- USE FOR N, USE FOR YEAR TOO?
      numericInput(inputId = "obs",
                   label = "Number of observations to view:",
                   value = 10)
    ),

    # panel for output
    mainPanel(

      # Output: Histogram ----
      plotOutput(outputId = "distPlot")

    )
  )
)

server <- function(input, output) {
  dataInput <- reactive({
  getSymbols(input$symb, src = "google", 
             from = input$dates[1],
             to = input$dates[2],
             auto.assign = FALSE)
})

output$plot <- renderPlot({
  chartSeries(dataInput(), theme = chartTheme("white"), 
              type = "line", log.scale = input$log, TA = NULL)
})
  
}

shinyApp(ui = ui, server = server)
```

0. **Who earned most?** Visualize the payroll information (total payment with breakdown into base pay, overtime pay, and other pay, Department, Job Title) of the top $n$ highest paid LA City employees in a specific year. User specifies $n$ (default 10) and year (default 2017).

0. **Which departments earn most?** Visualize the mean or median payroll, with breakdown into base pay, overtime pay, and other pay, of top $n$ earning departments. User specifies $n$ (default 5), year (default 2018), and method (mean or median, default median).

0. **Which departments cost most?** Visualize the total payroll, with breakdown into base pay, overtime pay, and other pay, of top $n$ expensive departments. User specifies $n$ (default 5) and year (default 2017).

0. Visualize any other information you are interested in.

0. Publish your Shiny app to <https://www.shinyapps.io> and share the link.

## Q2 LA City Parking War

The SQLite database `/home/m280data/la_parking/LA_Parking_Citations.sqlite` on teaching server contains information about parking tickets in LA City. It was downloaded from [LA Open Data Portal](https://data.lacity.org/A-Well-Run-City/Parking-Citations/wjz9-h9np). Connect to the database and answer following questions using plots and summary statistics. In this exercise, you are **not** allowed to load whole data into memory. Use the _transform in database, plot in R_ strategy.

```{bash}
ln -sf /home/m280data/nyc_parking/NYParkingViolations.csv NYParkingViolations.csv
```

```{r}
# reading in data and editing header names
library("tidyverse")
library("lubridate")
system.time({
  nyc <- read_csv("NYParkingViolations.csv") %>%
    setNames(str_replace_all(names(.)," ", "_")) %>%
    mutate(Issue_Date = mdy(Issue_Date)) %>% 
    mutate(Issue_Day = day(Issue_Date),
           Issue_Month = month(Issue_Date),
           Issue_Year = year(Issue_Date),
           Issue_WDay = wday(Issue_Date, label=TRUE))
})

# creating database
library("DBI")
library("RSQLite")
if (Sys.info()[["sysname"]] == "Linux") {
  db <- dbConnect(RSQLite::SQLite(), 
                  dbname = "/home/m280data/nyc_parking/NYParkingViolations.sqlite")
} else if (Sys.info()[["sysname"]] == "Darwin") {
  db <- dbConnect(RSQLite::SQLite(), 
                  dbname = "./NYParkingViolations.sqlite")
}
# dbWriteTable(db, "nyc", nyc, overwrite = TRUE)
dbListTables(db)
```

1. How many tickets are in this data set? Which time period do these tickets span? Which years have most data?

```{r}
# creating object from database
nyc_sql <- dplyr::tbl(db, "nyc")
range <- range(nyc$Issue_Date)
year <- nyc$Issue_Date %>% year() %>% table()
ticket <- nyc_sql %>% summarize(n = n())
```
There are 9,095,621 tickets in this dataset. These tickets span from 1970-12-18 to 2069-12-23. The year with the most tickets is 2014 with 4,716,512 entries. The year with the 2nd most is 2013 with 4,379,109 entries.

0. When (which hour, weekday, month day, and month) are you most likely to get a ticket and when are you least likely to get a ticket?

0. Which car makes received most citations?

```{r}
maker <- nyc_sql %>%
  group_by(Vehicle_Make) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
maker
```
Ford received the most citations at 1,248,019 entries.

0. How many different colors of cars were ticketed? Which color attracted most tickets?

```{r}
colorR <- unique(table(nyc$Vehicle_Color))
color <- nyc_sql %>%
  group_by(Vehicle_Color) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
colorR
color
```
White attracted the most tickets at 1,348,510 entries.

0. What are the most common ticket types?

```{r}
type <- length(unique(nyc$Violation_Post_Code))
type
# is this the right variable to calculate? why nyc and not nyc_sql? typos?
```

0. How much money was collected on parking tickets in 2016, 2017 and 2018?
```{r}
y2016 <- filter(nyc, Issue_Year == 2016)
m2016 <- y2016 %>%
  summarize(n = sum(y2016$Violation_Code))
m2016
# is this the right thing to be summing?
```

0. If you've been ticketed in LA County, did you find your ticket in this data set?

No, I have not been ticketed in LA County.

0. Read the blog <http://www.brettrics.com/9-million-parking-tickets-la/> and try to reproduce plots using your data.
